---
title: "Exploratory Data Analysis on Earthquake Data"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(sf)
library(dplyr)

```

# Plagiarism declaration

# Data Wrangling

The data from the "Earthquakes 1965 - 2016.csv" and "query.csv" files were read into R as dataframes. The dataframes were named `earthquakes` and `query`.
```{r}
earthquakes <-read.csv("Earthquakes 1965 - 2016.csv")
query <- read.csv("query.csv")
```

The column names and their respective data types for the `earthquakes`dataframe were viewed, they were as follows:
```{r}
str(earthquakes)
```

The column names and their respective data types for the `query` dataframe were viewed, they were as follows:
```{r}
str(query)
```
In the `query` table the `time` column contains milliseconds and is in a different format from that in the `earthquakes` table. The miliseconds were removed from the column.

```{r}
query <- query %>%
  mutate(time = ymd_hms(time)) # remove the milliseconds from the column
```
Rename time column to datetime
```{r}
#code here
```

```{r}

earthquakes <-read.csv("Earthquakes 1965 - 2016.csv", header = TRUE)

earthquakes$DateTime <- paste(earthquakes$Date, earthquakes$Time) # Combine the `Date` and `Time` columns using paste() function
earthquakes$DateTime <- as.POSIXct(earthquakes$DateTime, format="%m/%d/%Y %H:%M:%S") # convert `DateTime` column to a POSIXct (datetime) datatype
earthquakes$DateTime <- ymd_hms(earthquakes$DateTime) # convert to a better data format

str(earthquakes)

```

The `Date` and `Time` columns in the Earthquakes dataframe were merged into one column `DateTime`
```{r}
earthquakes$DateTime <- as.POSIXct(paste(earthquakes$Date, earthquakes$Time), format="%m/%d/%Y %H:%M:%S", tz="UTC")
```